The Need to Develop a Holistic Infectious Disease Surveillance System

### 2024-10-28 manuscript/manuscript.qmd
Add Matt's abstract suggestions

  Infectious disease surveillance and outbreak detection systems often utilize diagnostic testing to validate case identification.
  The metrics of sensitivity, specificity, and positive predictive value are commonly discussed when evaluating the performance of diagnostic tests, and to a lesser degree, the performance of outbreak detection systems.
  However, the interaction of the two levels’ (the test and the alert system) metrics, is typically overlooked.
  Here, we describe how equivalent regions of detection accuracy can exist over a range of diagnostic test characteristics, examining the sensitivity to background noise structure.

@@

  We generated a stochastic SEIR model with importation to simulate true measles and non-measles sources of febrile rash (noise) daily incidence.
  We simulated the noise to have in-phase, out-of-phase, non-seasonal dynamics, with respect to the measles simulation.
  We also examined noise where each day were independent Poisson draws.
  For rapid diagnostic test (RDT) and enzyme-linked immunosorbent assay (ELISA) like tests, we varied the reporting rate, and the observed daily incidence threshold required to trigger an alert, and computed the: percentages of alerts that were correct (analogous to the positive predictive value), the percentage of outbreaks detected (analogous to the sensitivity), the delay between outbreak initiation and alert, and the number of (un)avoidable outbreak cases.
  Summarizing the percentage of alerts correct and outbreaks detected into an accuracy metric, we selected the “optimal threshold” for each combination of diagnostic test and reporting rate, presenting the summary statistics above.

@@

  The optimal threshold for each test increased monotonically as the percentage of clinic visited who were tested increased, though there are adjacent reporting rates that have identical thresholds due to the integer requirement for the daily threshold.
  With Poisson-only noise, similar outbreak detection accuracies could be achieved with imperfect RDT-like tests as with ELISA-like diagnostic tests (c.
  93%), given moderately high testing rates.
  With larger delays (14 days) between the ELISA test procurement and result date, RDTs could outperform the ELISA.
  Similar numbers of unavoidable cases and outbreak alert delays could be achieved between the test types.
  With dynamical, in-phase, noise, however, the accuracy of ELISA scenarios was far superior to those achieved with RDTs (c.
  93% vs. 73%). For dynamical noise, RDT-based scenarios typically favored more sensitive alert threshold than ELISA-based scenarios (at a given reporting rate), observed with lower numbers of unavoidable cases and detection delays.

@@

  The performance of an outbreak detection system is highly sensitive to the structure and the level of background noise.
  Under the assumption that the noise is relatively static over time (as influenza-like illness was early in the COVID-19 pandemic), RDTs can perform as well as ELISA in a surveillance system.
  However, when the noise is non-stationary, imperfect tests cannot overcome their accuracy limitations through higher testing rates. 
toc: false
tbl-cap-location: bottom
format:
  generic-manuscript-typst:
    keep-typ: true
jupyter: julia-1.10
---

>>>>

  Infectious disease surveillance and outbreak detection systems often utilize diagnostic testing to validate case identification.
  The metrics of sensitivity, specificity, and positive predictive value are commonly discussed when evaluating the performance of diagnostic tests, and to a lesser degree, the performance of outbreak detection systems.
  However, the interaction of the two levels’ (the test and the alert system) metrics, is typically overlooked.
  Here, we describe how equivalent regions of detection accuracy can exist over a range of diagnostic test characteristics, examining the sensitivity to background noise structure and magnitude.

@@

  We generated a stochastic SEIR model with importation to simulate true measles and non-measles sources of febrile rash (noise) daily incidence.
  We generated time series of febrile rash (i.e., measles clinical case definition) by summing the daily incidence of measles and either independent Poisson noise or non-measles dynamical noise (consistent with rubella virus).
  For each time series we assumed a fraction of all cases were seen at a healthcare clinic, and a subset of those were diagnostically confirmed using a test with sensitivity and specificity consistent with either a rapid diagnostic test (RDT) or an enzyme-linked immunosorbent assay (ELISA).
  From the resulting time series of test-positive cases, we define an outbreak alert as the exceedance of a threshold by the 7-day rolling average of observed (test positive) cases.
  For each threshold level, we calculated percentages of alerts that were aligned with an outbreak (analogous to the positive predictive value), the percentage of outbreaks detected (analogous to the sensitivity), and combined these two measures into an accuracy metric for outbreak detection.
  We selected the optimal threshold as the value that maximizes accuracy.
  We show how the optimal threshold and resulting accuracy depend on the diagnostic test, reporting rate, and the type and magnitude of the non-measles noise.

@@

  The optimal threshold for each test increased monotonically as the percentage of clinic visits who were tested increased.
  With Poisson-only noise, similar outbreak detection accuracies could be achieved with imperfect RDT-like tests as with ELISA-like diagnostic tests (c.
  93%), given moderately high testing rates.
  With larger delays (14 days) between the ELISA test administration and result date, RDTs could outperform the ELISA.
  Similar numbers of unavoidable cases and outbreak alert delays could be achieved between the test types.
  With dynamical noise, however, the accuracy of ELISA scenarios was far superior to those achieved with RDTs (c. 93% vs. 73%). For dynamical noise, RDT-based scenarios typically favored more sensitive alert threshold than ELISA-based scenarios (at a given reporting rate), observed with lower numbers of unavoidable cases and detection delays.

@@

  The performance of an outbreak detection system is highly sensitive to the structure and the magnitude of background noise.
  Under the assumption that the noise is relatively static over time, RDTs can perform as well as ELISA in a surveillance system.
  However, when the noise is temporally correlated, as from a separate SEIR process, imperfect tests cannot overcome their accuracy limitations through higher testing rates. 
toc: false
tbl-cap-location: bottom
format:
  generic-manuscript-typst:
    keep-typ: true
jupyter: julia-1.10
---
